Evolution of pre-copulatory and post-copulatory inbreeding behaviour and associated polyandry

### 2016-11-01 ms_pre_post.Rmd
Edit through Model section, update bib to include recent Duthie and Reid AmNat paper

Previous models implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., Duthie and Reid *In press*; @Parker1979; @Parker2006; @Duthie]. Such models thereby isolate the invasion fitness of this single strategy. However, when both pre- and post-copulatory strategies can affect realised inbreeding, it cannot be assumed that both strategies will invade simultaneously, nor that invasion fitness of one strategy will be independent of the other For example, if pre-adaptation or rapid selection results in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect of such invading alleles could be weak or negligible.

@@

To analyse our model, we track mean values of alleles underlying tendency for polyandry ($P_{a}$), pre-copulatory inbreed strategy ($M_{a}$), and post-copulatory inbreeding strategy ($F_{a}$) in a focal population. We present evolutionary dynamics of these mean allele values across different parameter combinations to test our three focal hypotheses.

To test our hypothesis that selection for polyandry is driven by a female's ability to bias fertilisation toward inbreeding avoidance through post-copulatory mechanisms, we compare simulations in which all three phenotypes including tendency for polyandry ($P_{p}$), pre-copulatory inbreeding strategy ($M_{p}$), and post-copulatory strategy ($F_{p}$) evolve, with an identical set of simulations in which $F_{p}$ cannot evolve given $c_{P} = \{0, 0.0025, 0.005,  0.01\}$. To stop $F_{p}$ from evolving, we sever the connection from $F_{a}$ to $F_{p}$ such that all $F_{g}$ genotypes resulting from $F_{a}$ cause random fertilisation with respect to kinship (i.e, $F_{p}=0$). Hence $F_{a}$ alleles have no phenotypic effect and polyandry cannot cause polyandrous females to benefit through an increased ability to bias fertilisation toward inbreeding avoidance. By comparing mean $P_{a}$ values over generations when post-copulatory inbreeding avoidance also evolves versus when fertilisation is random, we test whether or not evolution of post-copulatory inbreeding strategy increases selection for polyandry.

To test the hypothesis that selection for a costly strategy of inbreeding avoidance will be negligible given a less costly alternative strategy, we quantify change in $M_{a}$ and $F_{a}$ over generations in simulations where the cost of pre-copulatory inbreeding strategy $c_{M}=0$ but the cost of post-copulatory inbreeding strategy $c_{F}=0.02$, and where $c_{F}=0$ but $c_{M}=0.02$. We contrast these simulations with evolution of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). A contrast between the baseline expectation for how costs affect evolution of a focal inbreeding strategy in the absence of an alternative strategy, and evolution of a focal strategy given that an alternative strategy also evolves, tests whether or not a relatively low cost alternative strategy decreases selection for a focal strategy.

To test the hypothesis that the existence of an adaptive inbreeding strategy will decrease the strength of selection on, and thereby preclude the invasion of, a complementary strategy, we first used exploratory simulations to determine evolution of pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy and associated polyandry, in isolation. To test whether or not pre-copulatory inbreeding avoidance was predicted to evolve when adaptive polyandry to avoid inbreeding was already established, we initiated pre-copulatory inbreeding strategy allele values at zero ($M_{a}=0$), but fixed post-copulatory inbreeding allele values at $F_{a}=-10$ and polyandry allele values at $P_{a}=1$ (i.e., $F_{p}$ and $P_{p}$ did not evolve). Similarly, to test whether or not post-copulatory inbreeding strategy evolved to avoid inbreeding, we initiated post-copulatory inbreeding strategy allele values and polyandry allele values at zero ($F_{a}=0$; $P_{a}=0$), but fixed pre-copulatory inbreeding strategy allele values at $M_{a}=-10$. For reference, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full sibling and $13.5$ times times less likely to choose a first cousin than a non-relative in pre- and post-copulatory choice, respectively.

In all simulations, we record mean values of $P_{a}$, $M_{a}$, $F_{a}$ in each generation. Each combination of parameter values simulated is replicated $40$ times, and grand mean values and standard error of means is calculated in each generation across replicate simulations. These analyses allow us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on post-copulatory inbreeding strategy might be weak if polyandry is rare). For all replicates, we set the maximum number of generations to 40000, which exploratory simulations and previous modelling [@Duthie] showed to be a sufficient number of generations for inferring long-term dynamics of mean allele values and therefore selection on phenotypes.  For all replicates, we set $\rho=5$ immigrants, which produced a range of kin and non-kin in each generation allowing females could express inbreeding strategies, and $n=8$ offspring, which was sufficient to keep populations consistently at carrying capacities and avoid population extinction. Values of $K_{f}$ and $K_{m}$ were both set to 100 because previous modelling has shown that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but populations are also not so small that selection is typically overwhelmed by genetic drift (Duthie and Reid *in press*).

>>>>

Previous models implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., @Parker1979; @Parker2006; @Duthie; @Duthie2016a]. Such models thereby isolate the invasion fitness of this single strategy. However, when both pre- and post-copulatory strategies can affect realised inbreeding, it cannot be assumed that both strategies will invade simultaneously, nor that invasion fitness of one strategy will be independent of the other For example, if pre-adaptation or rapid selection results in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect of such invading alleles could be weak or negligible.

@@

To test the hypothesis that evolution of post-copulatory inbreeding avoidance can facilitate evolution of costly polyandry, we compare simulations in which polyandry and pre-copulatory and post-copulatory inbreeding can all evolve, with otherwise identical simulations in which post-copulatory inbreeding strategy cannot evolve given four different costs of polyandry ($c_{P} = \{0, 0.0025, 0.005,  0.01\}$). To prevent $F_{p}$ from evolving, we sever the connection from $F_{a}$ to $F_{p}$ such that all $F_{g}$ genotypes cause random fertilisation with respect to kinship. Hence $F_{a}$ alleles have no phenotypic effect, meaning that polyandrous females cannot enact post-copulatory inbreeding avoidance.

To test the hypothesis that selection for a costly strategy of inbreeding avoidance will be negligible given a less costly alternative strategy, we quantify change in $M_{a}$ and $F_{a}$ over generations in simulations where the cost of pre-copulatory inbreeding strategy $c_{M}=0$ but the cost of post-copulatory inbreeding strategy $c_{F}=0.02$, and vice versa. We compare these simulations with evolution of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$).

To test the hypothesis that the existence of adaptive post-copulatory inbreeding avoidance will preclude the invasion of pre-copulatory inbreeding avoidance, or vice versa, we first used exploratory simulations to determine evolution of pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy and associated polyandry, in isolation. To test whether pre-copulatory inbreeding avoidance would evolve when adaptive polyandry and post-copulatory inbreeding avoidance was already established, we initiated $M_{a}$ allele values at zero, but fixed $F_{a}$ allele values at $-10$ and $P_{a}$ allele values at $1$ (i.e., meaning that $F_{p}$ and $P_{p}$ did not evolve). Similarly, to test whether post-copulatory inbreeding avoidance evolved when pre-copulatory inbreeding avoidance was already established, we initiated $F_{a}$ and $P_{a}$ allele values at zero but fixed $M_{a}$ allele values at $-10$. Consequently, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full brother and $13.5$ times times less likely to choose a first cousin than a non-relative in pre-copulatory and post-copulatory choice, respectively.

In all simulations, we record mean values of $P_{a}$, $M_{a}$, $F_{a}$ in each generation. Each combination of parameter values simulated is replicated $40$ times, and grand mean values and standard errors of means are calculated in each generation across replicates. These analyses allow us to infer how allele values change over generations in response to costs, but also in response to the changing values of other alleles and therefore potentially interacting phenotypes (e.g., selection on post-copulatory inbreeding strategy might be weak if polyandry is rare). For all replicates, we set the maximum number of generations to 40000, which exploratory simulations and previous modelling [@Duthie] showed to be a sufficient number of generations for inferring long-term dynamics of mean allele values and therefore selection on phenotypes.  For all replicates, we set $\rho=5$ immigrants, which produced a range of kin and non-kin in each generation allowing females to express inbreeding strategies, and $n=8$ offspring, which was sufficient to keep populations consistently at carrying capacities and avoid population extinction. Values of $K_{f}$ and $K_{m}$ were both set to 100 because previous modelling has shown that populations of this size are small enough that mate encounters between kin occur with sufficient frequency for selection on inbreeding strategy, but populations are also not so small that selection is typically overwhelmed by genetic drift [@Duthie2016a].
